Split-Apply-Combine with Dynamic Grouping

### 2023-03-03 paper/article.Rnw
wip

\Abstract{
foo
}

@@

\section{Introduction}
The operation of splitting a data set into non-overlapping groups, computing an
aggregate for each group, and combining the results into a new dataset is one
of the most common operations in data analyses. Indeed, any software for data
analyses includes some functionality for this. For example, the combination of
\code{split}/\code{lapply}/\code{unsplit} as well as \code{aggregate} have been
a part of the \proglang{S} \citep{becker1988new} and \proglang{R} \citep{rcore}
languages for a long time. For \proglang{R} there are several packages that
implement functionality for this, including \pkg{plyr}
\citep{wickham2011split}, it's successor \pkg{dplyr} \citep{wickham2022dplyr},
it's drop-in replacement \pkg{poorman} \citep{eastwood2022poorman}, and
\proglang{R} packages \pkg{collapse} \citep{krantz2022collapse} and
\pkg{data.table} \citep{dowle2022datatable}. In \proglang{Python} the
\code{pandas} package implements several methods for grouping records and
aggregating over one or more columns in data frame objects. Similarly, the
\pkg{DataFrames} package for \proglang{julia} implements such functionality
\citep{kaminski2022dataframes}.

In all packages mentioned, the calculation for each group uses data available
within the group. However, there are valid use cases where a group aggregate is
determined using attributes from out-of-group entities. One example where this
occurs is in the area of small area estimation (see e.g.
\citet{molina2015sae}. Here, one wishes to estimate an aggregate for a group,
for example a geographical region, or a detailed population subset, where the
number of (sampled) observations so small that the variance of the estimate
would be unacceptably large. In small area estimation (SAE) one trades bias for
variance by `borrowing statistical strength' from out-of-group records. The
out-of-group records can be obtained, for example by combining the original
small group with a group of records that are deemed similar in certain
respects. A second area where out-of-group records play a role is in certain
hot-deck imputation methods \citep{andridge2010review}.  In
$k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
records that are preferably in the same group, but this condition may be
relaxed if there are not enough records in the group.  In the \pkg{VIM} package
for \proglang{R} \citep{kowarik2016imputation}, this is controlled by a
combination of the Gower distance and setting conditions on minimal number of
donors. In practice, imputation is often performed via a fall-though scenario,
where one first tries to estimate a model within a group, but if the group is
too small for a reliable estimate of model parameters, the group is enlarged by
combining similar groups, similar to the small-area estimation scenario. A
third example comes from the field of statistical disclosure control (SDC)
\citep{hundepool2012statistical}. In this field, the so-called $p$\% rule
states that the value in a grouped aggregation may not be determined for more
than $p$ percent by a single entity. This occurs for example, when a certain
economic activity in a geographic region is dominated by a single company. In
such cases the value can either not be published (suppression), or one
publishes an aggregate over larger groups. Similarly for microdata there is the
concept of $k$-anonymity, which states that each combination of properties must
occur at least $k$ times in a data set so as to prevent reidentification or
disclosure of individual properties. One way to achieve this is to combine
detailed groupings based on categorical variables into larger groups. For
example by combining groups that in a hierarchical classification have the same
parent.

Both SAE and SDC are supported by \proglang{R} and other free software. Small
area estimation methodology has for example been implemented in the \pkg{sae}
package of \cite{molina2015sae} and the \pkg{hbsae} package of
\cite{boonstra2022hbsae}.  Disclosure control methodology is implemented in the
free and open source standalone software $\tau-$Argus for tabular data and
$\mu$-Argus for microdata \citep{argus2022}, and is also supported by
\proglang{R} packages \pkg{sdcMicro} and \pkg{sdcTable}
\citep{templ2015statistical, meindl2022sdctable}. Regarding imputation
methodology, the CRAN task view on missing
data\footnote{https://cran.r-project.org/web/views/MissingData.html} currently
lists 203 \pkg{R} packages that support some form of estimating missing data.
The \pkg{simputation} package \citep{loo2022simputation} seems to be the only
one that allows for some kind of fall-through scenario for selecting methods,
but it does not allow for dynamic grouping.

@@

The \code{accumulate} package serves the use case where a user wishes to
compute aggregates for a certain grouping of records. However, if a certain
instance of a group does not meet user-defined quality requirements, the set of
records is expanded by (recursively) collapsing the grouping according to a
user-defined scheme. For example, given some financial data on companies, one
wishes to compute the average profit to turnover ratio for each combination of
economic activity and size class. If for a certain combination of economic
activity and size class there are too few records for a reliable estimate, one
could drop size class and compute the average ratio over all records within a
certain economic activity. Or, one could choose to course-grain economic
activities by collapsing groups with activities that are deemed similar enough.

The package has been developed with the following design choices in mind.
First, the interface should be easy to learn for \proglang{R} users, and thus
should resemble existing popular interfaces where possible. Second, users
should be free to define any possibly multi-step, collapsing scheme.  Here, we
keep in mind that collapsing schemes may be constructed manually based on
domain knowledge and that users may want to experiment with several schemes
before deciding on a final solution. This calls for a certain separation of
concerns between defining collapsing schemes and applying them to data.  The
package should also support collapsing schemes that follow naturally from
hierarchical classification systems. Third, users should have the flexibility
to define any quality requirement on the grouping while common quality
requirements are supported out of the box. Common quality requirements include a
minimum number of records, or a minimum fraction of records without missing
values, or a minimum number of records with non-zero values.

The rest of this paper is organized as follows. In the next section we formally
define the concepts of the quality requirements and a collapsing scheme. We
describe the fundamental algorithm in pseudocode and compare it with the
standard grouping scenario. We also discuss the time and space complexity and
some caveats with implementation and performance. This section is mainly of
interest for developers who wish to include such functionality in their own
package. In Section~\ref{sect:accumulate} we describe \pkg{accumulate} package
and demonstrate how the design choices have determined its implementation.  We
give some toy examples to extensively discuss the input and output of the main
functions, discuss how to define tests in some detail, and also explain how one
can program over the functions in the package.  Section~\ref{sect:example}
works through a realistic example based on a synthetic dataset of economic
microdata. We conclude with a summary and outlook in
Section~\ref{sect:conclusion}.

\section{Dynamic grouping} \label{sect:dynamicgrouping}
In this section the illustrate the concept of dynamic grouping with a minimal
worked example. This example is not very realistic, but it is constructed
to be simple enough so the whole procedure can be followed in detail.

@@

The output of both functions are of the same form. The columns of the output
data frame and can schematically be represented as follows. 
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate (0 indicating no collapse), and the last set of columns
store the aggregates.

Both functions support two different interfaces for specifying collapsing
schemes through the \code{collapse} parameter. The first and most general is
the formula interface, which requires that the collapsing sequence is
represented as variables in the data set to be aggregated. The second is the
tabular interface, where a table similar to~\eqref{eq:tablerep} is expected.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing linear models}
Generalizing further, one may be interested in estimating (statistical) models.
With the \code{cumulate} function it is possible to specify complex aggregates
that go beyond a simple number. Below, we estimate the following linear model
\begin{displaymath}
\code{total} = \beta_0 + \beta_i\code{industrial} + \varepsilon,
\end{displaymath}
demanding that there are at least 10 records where both predictor and predicted
variable are available.
<<>>=
r <- cumulate(producers
         , collapse = sbi*size ~ sbi + sbi2 + sbi1
         , test = min_complete(n=5, vars=c("total","industrial"))
         , model = lm(total ~ industrial)
         , mean  = mean(total,na.rm=TRUE))
head(r,3)
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formal description and algorithms}
In this section we give a formal description of the algorithm for aggregation
with dynamic grouping.  We start by giving an algorithm for ordinary
split-apply-combine to show where the algorithm must be generalized to allow
for a collapsing scheme.

@@

Comparing Algorithms~\ref{alg:sac} and~\ref{alg:saccg}, we see that the
standard split-apply-combine algorithm has worst-case runtime complexity
$O(|A|)$, which is equal to the best case $\Omega(|A|)$. The
split-apply-combine with collapsing groups algorithm has best case
$\Omega(|A|)$ equal to standard split-apply-combine but worst case $O(n|A|)$.
In fact, the best case for Algorithm~\ref{alg:saccg} is achieved by setting
$\beta:d\mapsto\texttt{True}$. It is not hard to see that in this case
Algorithm~\ref{alg:saccg} reduces to Algorithm~\ref{alg:sac}. The worst case is
achieved by setting $\beta: d\mapsto \texttt{False}$. In that case the while
loop in Line~\ref{line:while} is iterated $n-1$ times (yielding total $n$
executions of $f^*$) while Line~\ref{line:R} is never executed.

\section{Summary and conclusion}\label{sect:conclusion}

\begin{itemize}
\item{Record-wise results}
\item{Record-wise imputation}
\item{Representation of hierarchical classifications: parent-child lists vs tabular}
\end{itemize}

\newpage
\bibliography{refs}

>>>>

\Abstract{
Partitioning a data set by one or more of its attributes and computing an
aggregate for each part is one of the most common operations in data analyses.
There are use cases where the partitioning is determined dynamically by
collapsing smaller subsets into larger ones, to ensure sufficient support for
the computed aggregate. These use cases are not supported by software
implementing split-apply-combine types of operations. This paper presents the
\proglang{R} package \code{accumulate} that offers convenient interfaces for
defining grouped aggregation where the grouping itself is dynamically
determined, based on user-defined conditions on subsets, and a user-defined
subset collapsing scheme. Finally, we demonstrate and analyze the pseudocode
underlying this method.
}

@@

\section{Introduction}
The operation of splitting a data set into non-overlapping groups, computing an
aggregate for each group, and combining the results into a new dataset is one
of the most common operations in data analyses. Indeed, any software for data
analyses includes some functionality for this. For example, the combination of
\code{split}/\code{lapply}/\code{unsplit} as well as \code{aggregate} have been
a part of the \proglang{S} \citep{becker1988new} and \proglang{R} \citep{rcore}
languages for a long time. For \proglang{R} there are several packages that
implement functionality for this, including \pkg{plyr}
\citep{wickham2011split}, it's successor \pkg{dplyr} \citep{wickham2022dplyr},
it's drop-in replacement \pkg{poorman} \citep{eastwood2022poorman}, and
performance-focused \proglang{R} packages \pkg{collapse}
\citep{krantz2022collapse} and \pkg{data.table} \citep{dowle2022datatable}. In
\proglang{Python} the \code{pandas} package implements several methods for
grouping records and aggregating over one or more columns in data frame
objects. Similarly, the \pkg{DataFrames} package for \proglang{julia}
implements such functionality \citep{kaminski2022dataframes}.

In all packages mentioned, the calculation for each group uses data available
within the group. However, there are valid use cases where a group aggregate is
determined using attributes from out-of-group entities. One example where this
occurs is in the area of Small Area Estimation (SAE, see \emph{e.g.}
\citet{rao2015small, molina2015sae}). Here, one wishes to estimate an aggregate
for a group, for example a geographical region, or a detailed population
subset, where the number of (sampled) observations so small that the variance
of the estimate would be unacceptably large. In small area estimation (SAE) one
trades bias for variance by `borrowing statistical strength' from out-of-group
records. The out-of-group records can be obtained, for example by combining the
original small group with a group of records that are deemed similar in certain
respects. A second area where out-of-group records play a role is in certain
hot-deck imputation methods \citep{andridge2010review}.  In
$k$-nearest-neighbours imputation for example, one finds a set of $k$ donor
records that are preferably in the same group, but this condition may be
relaxed if there are not enough records in the group.  In the \pkg{VIM} package
for \proglang{R} \citep{kowarik2016imputation}, this is controlled by a
combination of the Gower distance and setting conditions on minimal number of
donors. In practice, imputation is often performed via a fall-though scenario,
where one first tries to estimate a model within a group, but if the group is
too small for a reliable estimate of model parameters, the group is enlarged by
combining similar groups, similar to the small-area estimation scenario.

SAE as a special case is well supported by \proglang{R} and other free
software. Methodology has for example been implemented in the \pkg{sae} package
of \cite{molina2015sae} and the \pkg{hbsae} package of
\cite{boonstra2022hbsae}.  Regarding imputation methodology, the CRAN task view
on missing data\footnote{https://cran.r-project.org/web/views/MissingData.html}
currently lists 203 \pkg{R} packages that support some form of estimating
missing data.  The \pkg{simputation} package \citep{loo2022simputation} seems
to be the only one that allows for some kind of fall-through scenario for
selecting methods, but it does not allow for dynamic grouping.

@@

The \code{accumulate} package serves the use case where a user wishes to
compute aggregates for a certain grouping of records. However, if a certain
instance of a group does not meet user-defined quality requirements the set of
records is expanded by (recursively) collapsing the grouping according to a
user-defined scheme. For example, given some financial data on companies, one
wishes to compute the average profit to turnover ratio for each combination of
economic activity and size class. If for a certain combination of economic
activity and size class there are too few records for a reliable estimate, one
could drop size class and compute the average ratio over all records within a
certain economic activity. Or, one could choose to course-grain economic
activities by collapsing groups with activities that are deemed similar enough.

The package has been developed with the following design choices in mind.
First, the interface should be easy to learn for \proglang{R} users, and thus
should resemble existing popular interfaces where possible. Second, users
should be free to define any possibly multi-step, collapsing scheme. Here, we
keep in mind that collapsing schemes may be constructed manually based on
domain knowledge and that users may want to experiment with several schemes
before deciding on a final solution. This calls for a certain separation of
concerns between defining collapsing schemes and applying them to data.  The
package should also support collapsing schemes that follow naturally from
hierarchical classification systems. Third, users should have the flexibility
to define any quality requirement on the grouping while common quality
requirements are supported out of the box. Common quality requirements include
a minimum number of records, or a minimum fraction of records without missing
values, or a minimum number of records with non-zero values. Finally, the
package should support simple outputs such as counts or averages, but also
compound objects such as the output of model estimates.

The rest of this paper is organized as follows.  In the next section we
visually explain dynamic grouping via a collapsing scheme and introduce the
running example that will be used in Section~\ref{sect:accumulate} which
introduces the \pkg{accumulate} package, the main interfaces and helper
functions. In Section~\ref{sect:accumulate} we also discuss the common case of
collapsing via a predifined hiearchical classification scheme.
Section~\ref{sect:example} demonstrates the package on a realistic synthetic
dataset. The case of complex outputs is also demonstrated in this Section.  In
Section~\ref{sect:algorithm} we derive the pseudocode that solves the problem
of aggregating with dynamically collapsing groups. We show that it is a precise
and straightforward generalization of the standard split-apply-combine problem
and analyze its time complexity.

We describe the fundamental algorithm in pseudocode and compare it with the
standard grouping scenario. We also discuss the time and space complexity and

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Dynamic grouping} \label{sect:dynamicgrouping}
In this section the illustrate the concept of dynamic grouping with a minimal
worked example. This example is not very realistic, but it is constructed
to be simple enough so the whole procedure can be followed in detail.

@@

The output of both functions are of the same form. The columns of the output
data frame and can schematically be represented as follows. 
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate (0 indicating no collapse), and the last set of columns
store the aggregates. Output aggregates may be of a simple data type
(\code{numeric}, \code{character}, \code{logical},$\ldots$) or of a composed
type such as the output of a linear model. The latter case is demonstrated in
Section~\ref{sect:complex}.

Both functions support two different interfaces for specifying collapsing
schemes through the \code{collapse} parameter. The first and most general is
the formula interface, which requires that the collapsing sequence is
represented as variables in the data set to be aggregated. The second is a
tabular interface where each row starts with the one of the lowest-level group
labels, and subsequent columns contain labels of courser groups.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing complex aggregates} \label{sect:complex}
Generalizing further, one may be interested in estimating (statistical) models.
With the \code{cumulate} function it is possible to specify complex aggregates
that go beyond a simple number. Below, we estimate the following linear model
\begin{displaymath}
\texttt{total} = \beta_0 + \beta_i\texttt{industrial} + \varepsilon,
\end{displaymath}
demanding that there are at least 10 records where both predictor and predicted
variable are available.
<<>>=
r <- cumulate(producers
         , collapse = sbi*size ~ sbi + sbi2 + sbi1
         , test = min_complete(n=10, vars=c("total","industrial"))
         , model = lm(total ~ industrial))
head(r,3)
@
Here, the last column is a list of class \code{object_list}, where each
element is either an object of class \code{lm} or \code{NA}. Variables of class
\code{object_list} only differ from from a standard \proglang{R} list by their
print method.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formal description and algorithms} \label{sect:algorithm}
In this section we give a formal description of the algorithm for aggregation
with dynamic grouping.  We start by giving an algorithm for ordinary
split-apply-combine to show where the algorithm must be generalized to allow
for a collapsing scheme.

@@

Comparing Algorithms~\ref{alg:sac} and~\ref{alg:saccg}, we see that the
standard split-apply-combine algorithm has worst-case runtime complexity
$O(|A|)$, as determined by counting applications of $f^*$. This is equal to the
algorithm's best case $\Omega(|A|)$. The split-apply-combine with collapsing
groups algorithm also has best case $\Omega(|A|)$ but has worst case $O(n|A|)$
(with $n$ the number of collapsing steps). In fact, the best case for
Algorithm~\ref{alg:saccg} is achieved by setting $\beta:d\mapsto\texttt{True}$.
It is not hard to see that in this case Algorithm~\ref{alg:saccg} reduces to
Algorithm~\ref{alg:sac}. The worst case is achieved by setting $\beta: d\mapsto
\texttt{False}$. In that case the while loop in Line~\ref{line:while} is
iterated $n-1$ times (yielding total $n$ executions of $f^*$) while
Line~\ref{line:R} is never executed.

The analyses above leaves open the question of how the runtime depends on
application of the pullback $f^*$. First note that there is no difference in
applying the $f^*$ or $f^*\circ F_i^*\circ F_i$: in practice a nested function
call is not necessary as the collapsing scheme is fully available as input. The
pullback performs a lookup of records based on matching one or more attributes
with a set of attribute values. The time complexity of such operations are
typically reduced by proper preparation of the dataset. For example, in
databases one may define indices that speed up certain often-used lookup
operations (that then may rely on those indices). In the implementation of the
\code{accumulate()} package the pullback is (currently) implemented using
standard join operations as implemented by \proglang{R}'s standard
\code{merge()} function.  This implies that the time complexity of
\code{accumulate()} and \code{cumulate()} should be multiplied with the time
dependence of \code{merge()}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary and conclusion}\label{sect:conclusion}
The \proglang{R} package \code{accumulate} introduced in this paper offers
convenient interfaces for computing grouped aggregates where the grouping is
dynamically determined based on a user-defined conditions and a user-defined
collapsing scheme. We demonstrated that these interfaces can be used in
situations including multiple grouping variables and complex collapsing
schemes, as well as situations where multiple variables need to be aggregated
over. Moreover, the package supports the collection of complex aggregates, such
as model outputs and offers several service functions that aim to facilitate
the definition of the collapsing conditions.

Finally, the pseudocode underlying the package's main functions has been
formally analyzed and it is shown that aggregation with collapsing groups can
be interpreted as a generalization of standard grouped aggregation in a very
precise sense. We hope that this stimulates broader implementation of this
algorithm in software offering grouped aggregation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\bibliography{refs}
